---
title: "Practical Machine Learning Course Project"
author: "__Joyvalerie Mondejar__"
date: _Last Updated:_ `r format(Sys.time(), '%B %d, %Y')`
output: 
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
editor_options: 
  chunk_output_type: console
---

_Date Created_: `r format(as.Date("2020-01-14"), "%B %d, %Y")`


## Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about
personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of
enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their
behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular
activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from
accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts
correctly and incorrectly in 5 different ways. More information is available from the website here:
<http://groupware.les.inf.puc-rio.br/har> (see the section on the Weight Lifting Exercise Dataset). 


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
suppressMessages(library(caret))
suppressMessages(library(randomForest))
library(rpart)
suppressMessages(library(rpart.plot))
suppressMessages(library(data.table))
library(magrittr)
suppressMessages(library(rattle))      # used to plot the classification trees
library(here)
```

## Getting and Cleaning Data

The training data for this project are available here: <https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv>

The test data are available here: <https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv>

The data for this project come from this source: <http://groupware.les.inf.puc-rio.br/har>.

Loading data

```{r}
trainingSet <- read.csv("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv")
testingSet <- read.csv("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv")
```

Cleaning Data

```{r}
# see error proportion
NAProportion <- round(colMeans(is.na(trainingSet)), 2)
head(NAProportion, 20)
table(NAProportion)
# 93 columns have complete data, while 67 columns have 98% missing data
```

```{r}
# find index of the complete columns minus the first 7 columns since they seems (administrative data) to be not useful in predicting the outcome variable classe 
index <- which(NAProportion == 0)[-c(1:7)]
# subset the data
trainData <- trainingSet[, index]
testData <- testingSet[, index]
dim(trainData)
dim(testData)
```

```{r}
remove_cols <- nearZeroVar(trainData)     # removed the 33 columns with near zero variance
trainData <- trainData[, -remove_cols]
testData <- testData[, -remove_cols]
dim(trainData)
dim(testData)
trainData$classe <- as.factor(trainData$classe)
```

## Cross Validation

We set `testData` aside and split the `trainData` into two parts. We will allocate 70% of the `trainData` to train the model on and 30% to validate it.

```{r}
set.seed(3475)
inTrain <- createDataPartition(y = trainData$classe, p = 0.7, list = FALSE)
trainData <- trainData[inTrain, ]
validData <- trainData[-inTrain, ]
rbind(trainData = dim(trainData), validData = dim(validData), test = dim(testData))
```

We will now train the data using the remaining 52 variables.

## Model Building

### Prediction with Classication Trees

```{r decision-tree, cache=TRUE}
decisionTreeMod <- train(classe ~ ., data = trainData, method = "rpart")  # constructs trees based on the outcome and predictors
print(decisionTreeMod$finalModel)  # returns text summary of all nodes/spits in the constructed tree
```

```{r}
# Plots the classification tree with all nodes/splits
fancyRpartPlot(decisionTreeMod$finalModel)   # from rattle package
```

```{r}
# use the model to predict on validation dataset
predictTreeMod <- predict(decisionTreeMod, newdata = validData)
# confusion matrix of predicted results
cmtree <- confusionMatrix(predictTreeMod, validData$classe)
cmtree
```

```{r}
# plot matrix results
plot(cmtree$table, col = cmtree$byClass, 
     main = paste("Tree - Accuracy =", round(cmtree$overall['Accuracy'], 4)))
```

The decision tree has accuracy of 0.4998 and therefore the out-of-sample error 0.5002

### Prediction using Random Forest

```{r rf, cache=TRUE}
# Fitting a random forest model
set.seed(3475)
controlRF <- trainControl(method = "cv", number = 10, verboseIter = FALSE)
t1 <- Sys.time()
rfMod <- train(classe ~ ., data = trainData, method = "rf", prox = TRUE, trControl = controlRF)
rfMod
t2 <- Sys.time()
t2 - t1
rfMod$finalModel
```

```{r}
getTrainPerf(rfMod)
```

```{r}
# predcition on testData
predictRfMod <- predict(rfMod, newdata = validData)
cmRf <-  confusionMatrix(predictRfMod, validData$classe)
cmRf
```

The random forest model is better performing with accuracy of 1 and therefore the out-of-sample error rate is 0

```{r}
# plot matrix results
plot(cmRf$table, col = cmRf$byClass, 
     main = paste("Random Forest - Accuracy =",
                  round(cmRf$overall['Accuracy'], 4)))
```

### Prediction using Generalized Boosted Model

```{r gbm, cache=TRUE}
set.seed(3475)
controlGBM <- trainControl(method = "repeatedcv", number = 5, repeats = 1)
t1 <- Sys.time()
gbMod <- train(classe ~ ., data = trainData, method = "gbm", trControl = controlGBM, verbose = FALSE)
t2 <- Sys.time()
t2 - t1
gbMod$finalModel
```

```{r}
# prediction on validData
predictgbMod <- predict(gbMod, newdata = validData)
cmGb <- confusionMatrix(predictgbMod, validData$classe)
cmGb
```

The generalized boosting model has accuracy of 0.98 and therefore the out-of-sample error is 0.02

```{r}
# plot matrix results
plot(cmGb$table, col = cmGb$byClass, 
     main = paste("GBM - Accuracy =", round(cmGb$overall['Accuracy'], 4)))
```

## Decision

The accuracy of three classification models above are:  
a. Decision Tree: 0.4998  
b. Random Forest: 1  
c. Generalized Boosted Model: ~0.98

So, considering the model with the highest accuracy the Ramdom Forest model will be applied to predict the 20 different test cases.

```{r}
predictTest <- predict(rfMod, newdata = testData)
predictTest
```



